Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2016-05-02 doc/invt.tex
invt.tex: small changes
invt.tex: p/rho separation

\title{Optimal updating factor in adaptive flat-distribution simulations}

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for adaptive flat-distribution sampling methods,
  such as the Wang-Landau (WL) algorithm and metadynamics.
  %
  These sampling methods
  construct a bias potential to offset
  the potential of mean force
  by incremental updating loops.
  %
  The updating magnitude should decrease over time
  to reduce the fluctuation of the bias potential,
  and the optimal schedule optimizes
  the rate of convergence.
  %
  We derive the optimal schedule from
  an equation of motion
  of a free particle with a position-dependent mass,
  %
  and show that
  the initial updating magnitude is ideally half of
  the previous equilibrium value.
  %
  Further,
  we show that the WL updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence,
  and the optimal schedule of these schemes
  is given by a modified inverse-time formula.
  %
\end{abstract}

@@

To proceed, we first observe that
the average
$$
\bar v = \sum_{i = 1}^n p_i \, v_i
$$
increases steadily over time:
%
\begin{equation}
\frac{ d \bar v } { d t }
=
\sum_{i = 1}^n p_i \dot v_i
=
\alpha(t) \sum_{i = 1}^n h_i(t) = \alpha(t).
\label{eq:dvbardt}
\end{equation}
%
However, we expect the difference %between $v_i$ and $\bar v$,
%
\begin{equation}
  x_i \equiv v_i - \bar v = v_i - \sum_{j = 1}^n p_j \, v_j,
  \label{eq:x_def}
\end{equation}
%
to be small in the asymptotic regime.
Expanding Eq. \eqref{eq:h_ave} in $x_i$ yields
%
\begin{align}
  \langle h_i(t) \rangle
  &\approx
  \frac{                 p_i \, e^{- x_i} }
       { \sum_{ j = 1}^n p_j \, e^{- x_j} }
  \approx
  p_i \, (1 - x_i),
\label{eq:hdet_linearize}
\end{align}
%
where we have used $\sum_{j=1}^n p_j = 1$,
and
%
\begin{equation}
  \sum_{i = 1}^n p_i \, x_i = 0,
  \label{eq:px_sum}
\end{equation}
which follows directly from Eq. \eqref{eq:x_def}.
\note{For \eqref{eq:hdet_linearize}. Following the second step
$$
\langle h_i(t) \rangle
\approx
\frac{                 p_i (1 - x_i) }
     { \sum_{ j = 1}^n p_j (1 - x_j) }
.
$$
}%
From Eqs.
\eqref{eq:vt_diffeq},
\eqref{eq:h_split},
\eqref{eq:dvbardt}-\eqref{eq:hdet_linearize},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \frac{ \zeta_i(t) } { p_i } \right].
  \label{eq:dxdt_WL}
\end{equation}
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:dvbardt}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left( \frac{ h_i } { p_i } - 1 \right)
  \\
  &\stackrel{\mathrm{Eq.\; \eqref{eq:h_split}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left( \frac{ \langle h_i \rangle } { p_i } + \frac{ \zeta_i } { p_i } - 1 \right)
  \stackrel{\mathrm{Eq.\; \eqref{eq:hdet_linearize}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \frac{ \zeta_i } { p_i } \right].
\end{aligned}
$$
}

The total error of the bias potential can be written in
terms of $x_i$,
\begin{equation}
  \Err
  =
  \sum_{i = 1}^n p_i \, \langle x_i^2 \rangle
  =
  \sum_{i = 1}^n p_i \, \left\langle (v_i - \bar v)^2 \right\rangle
  .
\label{eq:error_sum}
\end{equation}
%
Our aim is to find the $\alpha(t)$
that minimizes the error $E$
at the end of the simulation period $T$.
%
Fortunately, it suffices to consider
a simpler one-variable problem.

@@

To simplify the ensuing discussion,
we shall further impose the
detailed balance condition:
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}.
  \label{eq:w_detailedbalance}
\end{equation}
%
It follows that
\begin{equation}
  \sum_{i = 1}^n p_i \, w_{ij}
  =
  \sum_{i = 1}^n p_j \, w_{ji}
  = p_j,
  \label{eq:w_balance}
\end{equation}
%
i.e., $\mathbf p$ is a left eigenvector of
$\mathbf w$ with eigenvalue $1$.

By assuming Eq. \eqref{eq:w_detailedbalance},
we can diagonalize $\mathbf w$ with a set of
eigenvectors
%
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij}
  =
  \lambda_k \, \phi_{kj}
  ,
\label{eq:eig_w}
\end{equation}
%
satisfying the orthonormal conditions\cite{vankampen}:
%
\begin{align}
  \sum_{k = 1}^n
    \phi_{ki} \, \phi_{kj}
  &=
  \delta_{ij} \, p_i,
  \label{eq:eig_orthonormal_cols}
  \\
  \sum_{i = 1}^n
    \frac{ \phi_{ki} \, \phi_{li} }
         { p_i }
  &=
  \delta_{kl}
  .
\label{eq:eig_orthonormal_rows}
\end{align}

Particularly, there is always an eigenvector
corresponding to the target distribution, $\mathbf p$.
%
We shall label this eigenvector as the first one,
and
%
\begin{equation}
  \phi_{1i} = p_i,
\label{eq:eigenmode1}
\end{equation}
%
which satisfies Eq. \eqref{eq:eig_orthonormal_rows}
for $k = l = 1$.
%
The corresponding eigenvalue is $\lambda_1 = 1$
by Eq. \eqref{eq:w_balance}.
%
Further, for $k \ge 2$, we have
%
\begin{equation}
  \sum_{ i = 1 }^n \phi_{ki}
  =
  \sum_{ i = 1 }^n
    \frac{ \phi_{ki} \, \phi_{1i} }
         { p_i }
  =
  0
  .
\label{eq:ortho1}
\end{equation}

@@

We can now simplify Eq. \eqref{eq:vt_diffeq_mbin}.
%
By using Eq. \eqref{eq:w_balance},
we can show that
Eqs. \eqref{eq:dvbardt} and \eqref{eq:px_sum}
remain true, and
%
$$
\begin{aligned}
  \dot x_i(t)
  &= \alpha(t) \sum_{j=1}^n w_{ij}
  \left( \frac{ h_j(t) } { p_j }  - 1 \right)
  \\
  &\approx
  -\alpha(t) \sum_{j = 1}^n
  w_{ij} \left[ x_j(t) - \frac{\zeta_j (t)}{p_j} \right],
\end{aligned}
$$
where
we have expanded the right-hand side
in linear terms of $x_j$
on the second line.

\note{The proof of Eq. \eqref{eq:dvbardt},
$$
\begin{aligned}
  \dot{ \bar v }
  &=
  \sum_{i = 1}^n p_i \, \dot v_i
  %
  \stackrel{\mathrm{Eq.\; \eqref{eq:vt_diffeq_mbin}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  %
  \sum_{j = 1}^n \alpha(t) \, \frac{ h_j(t) } { p_j }
                 \sum_{ i = 1 }^{n} p_i \, w_{ij}
  \\
  &
  \stackrel{\mathrm{Eq.\; \eqref{eq:w_balance}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \sum_{j = 1}^n \alpha(t) \, h_j(t)
  =
  \alpha(t).
\end{aligned}
$$
Equation \eqref{eq:px_sum} follows directly from the definition
Eq. \eqref{eq:x_def}.
}

Now by diagonalizing the matrix $\mathbf w$
using the eigenvectors defined in Eqs.
\eqref{eq:eig_w}-\eqref{eq:eig_orthonormal_rows},
we get a set of decoupled equations
%
\begin{equation}
  \dot y_k(t)
  =
  -\alpha(t) \, \lambda_k \, [y_k(t) - \theta_k(t)],
\label{eq:yt_diffeq}
\end{equation}
%
where
\begin{align}
  y_k(t)
  &=
  \sum_{j=1}^n
    \phi_{kj} \, x_j(t)
  ,
  \label{eq:y_def}
  \\
  \theta_k(t)
  &=
  \sum_{j=1}^n
    \phi_{kj} \frac{ \zeta_j(t) }
                   { p_j        }
  .
  \label{eq:eta_def}
\end{align}

@@

The error function defined in Eq. \eqref{eq:error_sum}
can be rewritten in terms of the eigenmodes
\begin{align}
  \Err
  \equiv
  \sum_{i = 1}^n p_i \,
    \left\langle x_i^2 \right\rangle
  =
  \sum_{i, j, k=1}^n \phi_{ki} \, \phi_{kj} \,
    \left\langle x_i \, x_j \right\rangle
  =
  \sum_{k = 1}^n
    \left\langle
      y_k^2
    \right\rangle
  ,
\notag
\end{align}
%
where we have used Eq. \eqref{eq:eig_orthonormal_cols}.
%
From Eq. \eqref{eq:eigenmode1},
the first eigenmode, $y_1 = \sum_{j=1}^n p_j \, x_j$,
vanishes identically by Eq. \eqref{eq:px_sum}.
%
Thus, we can start the last sum %in Eq. \eqref{eq:y2_sum}
from $k = 2$,
\begin{equation}
  \Err
  =
  \sum_{k = 2}^n
    \left\langle
      y_k^2
    \right\rangle
  .
\label{eq:y2_sum}
\end{equation}

@@

    % xtics
    \draw[] (0.0, 0.1) -- (0.0, 0.0) node[below] {$0$};
    \draw[] (2.2, 0.1) -- (2.2, 0.0) node[below] {$q(T)$};
  \end{tikzpicture}
  %\makebox[\linewidth][c]{
  %  \includegraphics[angle=0, width=0.9\linewidth]{fig/massq.pdf}
  %}
  %\caption{
  %  \label{fig:massq}
  %  Geometric characterization of the optimal schedule,
  %  $\alpha(t)$,
  %  via the normalized mass distribution, $m(Q)$.
  %  %
  %  \note{The figure was produced by \texttt{doc/fig/massq.gp}.
  %  }%
  %}
\end{center}
\end{figure}

@@

Many commonly-used updating schemes
can be characterized by a rigid window
centered on the current bin.
%
We call these schemes homogeneous,
or translationally invariant.
%
Naturally, $\rho_i = 1/n$ in this case.
%
An example is the Gaussian updating scheme
used in metadynamics.
%
Homogeneous updating schemes
share the same set of eigenvectors
but not the corresponding eigenvalues.
%
They have significantly fewer
parameters than the general case,
and the multiplication of the updating matrix,
$\mathbf w$,
can be reduced to a convolution with the window,
or the updating kernel.
%
Thus,
there is a one-to-one correspondence between
the eigenvalues of $\mathbf w$
and the updating kernel.

@@

\begin{table}[h]
  \caption{\label{tab:error_Gaussian}
    Error of the Gaussian updating scheme of $\sigma = 10$.
    $t_0$ is given by Eq. \eqref{eq:t0_sinc}.
    \note{
      The data were collected in
      \texttt{data/tab1} by running
      the program \texttt{invt}.
    }%
  }
  \setlength{\tabcolsep}{5pt}
  \begin{tabular} { l | p{2cm} p{2.5cm} p{1.8cm} }
    \hline
      Schedule
    &
      Inverse-time, Eq. \eqref{eq:alpha_invt1}
    &
      Modified inverse-time,
      Eq.~\eqref{eq:alpha_invtlambda}
      with optimized $\lambda$
    &
      Optimal,
      Eq. \eqref{eq:q_opt}
    \\
    \hline
    Global
    &
    $4.4(1) \times 10^{-6}$
    &
    $7.1 \times 10^{-7}$
    &
    $2.6(1) \times 10^{-7}$
    \\
    \hline
    Local
    &
    $3.8(2) \times 10^{-4}$
    &
    $1.8(1) \times 10^{-4}$
    &
    $1.0(1) \times 10^{-4}$
    \\
    \hline
  \end{tabular}
\end{table}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/xerr.pdf}
  }
  \caption{
    \label{fig:xerr}
    Error components for the Gaussian updating scheme
    with $\sigma = 10$
    from the optimal schedule
    and the inverse-time schedule, Eq. \eqref{eq:alpha_invt1}.
    %
    The latter schedule failed to damp out the errors
    from the long-wavelength (local) eigenmodes
    with $K \ge 5$.
    %
    For longer simulations,
    the error components from the optimal schedule
    would roughly maintain the same shape,
    but join the initial error curve
    at a lower level (not shown).
    %
    \note{This figure was produced by \texttt{doc/fig/xerr.gp}.
      The data were saved under the directory \texttt{data/xerr}.
    }%
  }
\end{center}
\end{figure}

@@

Here, we evaluate the integrals of
the autocorrelation functions,
$\Gamma_k$'s,
in two special cases
in Secs. \ref{sec:Gamma_perfect}
and \ref{sec:Gamma_onestep},
and describe a general method of measuring them
in Sec. \ref{sec:Gamma_measure}.

@@

Another special case is the
one-dimensional one-step process\cite{vankampen},
which models a local sampling process.
In each step, the system hops to
one of the two neighboring bins
with equal probability $1/2$.
%
The transition matrix is
%
\begin{equation}
  \arraycolsep=3.6pt\def\arraystretch{1.4}
  \mathbf A
  =
  \left(
    \begin{array}{cccccccc}
      \frac \sigma 2 & \frac 1 2 & 0 & \dots & \frac {1 - \sigma} 2 \\
      \frac 1 2 & 0         & \frac 1 2 & \dots & 0 \\
      \vdots & &  & & \vdots \\
      \frac {1 - \sigma} 2 & \dots &  & \frac 1 2 & \frac \sigma 2
    \end{array}
  \right),
\notag
%\label{eq:T_nn}
\end{equation}
%
where $\sigma = 0$ for a periodic variable, or $1$ otherwise.
%
For many updating schemes
(cf. Sec. \ref{sec:band-matrix}),
we can further assume that
the matrices $\mathbf A$ and $\mathbf w$
share the eigenvectors, but not the eigenvalues.
%
The eigenvalues of $\mathbf A$ are
$$
\Lambda_k
=
1 -
2 \sin^2 \frac{ (k - 1) \, g \, \pi }
              {       2 \, n        }
,
$$
where
$g = 2$ for a periodic variable, or $1$ otherwise.
%Note that these $\Lambda_k$'s are
%not to be confused with the eigenvalues of $\mathbf w$,
%$\lambda_k$'s.
%
In this case,
the autocorrelation function $\kappa_k(t)$
decays as $\kappa_k(0) \, \Lambda_k^t$,
with $\kappa_k(0) = 1$ being the same as
the value in perfect sampling.
%
Thus,
%$\Gamma_k$ is roughly twice the autocorrelation time, and
from Eq. \eqref{eq:Gamma_sum}, we get
%
\begin{equation}
  \Gamma_k
  =
  1 + 2 \, \sum_{t = 1}^\infty \Lambda_k^t
  =
  \frac{ 1 + \Lambda_k }
       { 1 - \Lambda_k }
  =
  \cot^2 \left[
    \frac{ (k - 1) \, g \, \pi }
         { 2 \, n }
  \right]
  .
\label{eq:Gamma_onestep}
\end{equation}

>>>>

\title{Optimal updating factor in adaptive flat-distribution-sampling simulations}

\begin{abstract}
  We present a method of computing the optimal schedule
  of the updating magnitude
  for adaptive flat-distribution-sampling methods,
  such as the Wang-Landau (WL) algorithm and metadynamics.
  %
  These sampling methods
  construct a bias potential to offset
  the potential of mean force
  by incremental updating loops.
  %
  The updating magnitude should decrease over time
  to reduce the fluctuation of the bias potential,
  and the optimal schedule optimizes
  the rate of convergence.
  %
  We derive the optimal schedule from
  an equation of motion
  of a free particle with a position-dependent mass,
  %
  and show that
  the initial updating magnitude is ideally half of
  the previous equilibrium value.
  %
  Further,
  we show that the WL updating scheme
  belongs to a class of bandpass updating schemes
  that are optimal for asymptotic convergence,
  and the optimal schedule of these schemes
  is given by a modified inverse-time formula.
  %
\end{abstract}

@@

To proceed, we first define an average\cite{
dama2014},
\begin{equation}
  \bar v(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$.
%
This distribution $\pmb \rho$
will be used to measure the error,
and is independent of
the target distribution $\mathbf p$.
%
For example,
in well-tempered metadynamics,
$\pmb\rho$ is always flat,
while $\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
barducci2008, dama2014}.
%
Then, we expect the difference %between $v_i$ and $\bar v$,
%
\begin{equation}
  x_i \equiv v_i - \bar v
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
to be small in the asymptotic regime.
%
It follows that
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i = 0
  ,
\label{eq:rhox_sum}
\end{equation}
and
\begin{equation}
  \frac{ d \bar v (t) }
       { d t }
  =
  \alpha(t)
  \sum_{ r = 1 }^n \frac{ h_r(t) } { p_r }
  .
\label{eq:dvbardt}
\end{equation}
%
Expanding Eq. \eqref{eq:h_ave} in $x_i$ yields
%
\begin{align}
  \frac{
    \langle h_i(t) \rangle
  }
  {
    p_i
  }
  &\approx
  \frac{                        e^{- x_i} }
       { \sum_{ j = 1}^n p_j \, e^{- x_j} }
  \approx
  1 - x_i + \sum_{r = 1}^n p_r \, x_r
  ,
\label{eq:hdet_linearize}
\end{align}
%
where we have used $\sum_{j=1}^n p_j = 1$.
%
\note{For \eqref{eq:hdet_linearize}. Following the second step
$$
\langle h_i(t) \rangle
\approx
\frac{                  p_i (1 - x_i) }
     { \sum_{ r = 1 }^n p_r (1 - x_r) }
=
\frac{                      p_i (1 - x_i) }
     { 1 - \sum_{ r = 1 }^n p_r \, x_r }
.
$$
}%
From Eqs.
\eqref{eq:vt_diffeq},
\eqref{eq:h_split},
\eqref{eq:x_def}-\eqref{eq:hdet_linearize},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
where
\begin{equation}
  \xi_i(t)
  =
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ r = 1 }^n
    \rho_r
    \frac{ \zeta_r(t) }
         {     p_r    }
  .
\label{eq:xi_def}
\end{equation}
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:dvbardt}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ h_r } { p_r }
  \right)
  \\
  &\stackrel{\mathrm{Eq.\; \eqref{eq:h_split}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ \langle h_r \rangle } { p_r }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:rhox_sum} \;and\; \eqref{eq:hdet_linearize}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

We define
the total error of the bias potential as
\begin{equation}
  \Err
  =
  \sum_{i = 1}^n \rho_i \, \langle x_i^2 \rangle
  =
  \sum_{i = 1}^n \rho_i \, \left\langle (v_i - \bar v)^2 \right\rangle
  .
\label{eq:error_sum}
\end{equation}
%
Our aim is to find the $\alpha(t)$
that minimizes the error $E$
at the end of the simulation period $T$.
%
Fortunately, it suffices to consider
a simpler one-variable problem.

@@

To simplify the ensuing discussion,
we shall further impose the
detailed balance condition:
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}.
  \label{eq:w_detailedbalance}
\end{equation}
%
for the distribution $\pmb \rho$
introduced in Eq. \eqref{eq:vbar_def}.
%
It follows that
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j,
  \label{eq:w_balance}
\end{equation}
%
i.e., $\mathbf \rho$ is a left eigenvector of
$\mathbf w$ with eigenvalue $1$.

By assuming Eq. \eqref{eq:w_detailedbalance},
we can diagonalize $\mathbf w$ with a set of
eigenvectors
%
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij}
  =
  \lambda_k \, \phi_{kj}
  ,
\label{eq:eig_w}
\end{equation}
%
satisfying the orthonormal conditions\cite{vankampen}:
%
\begin{align}
  \sum_{k = 1}^n
    \phi_{ki} \, \phi_{kj}
  &=
  \delta_{ij} \, \rho_i,
  \label{eq:eig_orthonormal_cols}
  \\
  \sum_{i = 1}^n
    \frac{ \phi_{ki} \, \phi_{li} }
         { \rho_i }
  &=
  \delta_{kl}
  .
\label{eq:eig_orthonormal_rows}
\end{align}

Particularly, there is always an eigenvector
corresponding to the distribution, $\pmb \rho$.
%
We shall label this eigenvector as the first one,
and
%
\begin{equation}
  \phi_{1i} = \rho_i,
\label{eq:eigenmode1}
\end{equation}
%
which satisfies Eq. \eqref{eq:eig_orthonormal_rows}
for $k = l = 1$.
%
The corresponding eigenvalue is $\lambda_1 = 1$
by Eq. \eqref{eq:w_balance}.
%
Further, for $k \ge 2$, we have
%
\begin{equation}
  \sum_{ i = 1 }^n \phi_{ki}
  =
  \sum_{ i = 1 }^n
    \frac{ \phi_{ki} \, \phi_{1i} }
         { \rho_i }
  =
  0
  .
\label{eq:ortho1}
\end{equation}

@@

We can now simplify Eq. \eqref{eq:vt_diffeq_mbin}.
%
By using Eqs. \eqref{eq:vbar_def}
and \eqref{eq:w_balance},
we can show that Eq. \eqref{eq:dvbardt}
remains true,
and by using Eqs.
\eqref{eq:x_def},
\eqref{eq:rhox_sum},
\eqref{eq:hdet_linearize},
\eqref{eq:xi_def},
and \eqref{eq:w_sumj}, we get
%
$$
\begin{aligned}
  \dot x_i(t)
  &= \alpha(t) \sum_{j=1}^n w_{ij}
  \left(
    \frac{ h_j(t) } { p_j }
    -
    \sum_{r = 1}^n
      \rho_r \frac{ h_r(t) } { p_r }
  \right)
  \\
  &\approx
  -\alpha(t) \sum_{j = 1}^n
  w_{ij} \left[ x_j(t) - \xi_j (t) \right]
  .
\end{aligned}
$$
%where
%we have expanded the right-hand side
%in linear terms of $x_j$
%on the second line.

\note{The proof of Eq. \eqref{eq:dvbardt},
$$
\begin{aligned}
  \dot{ \bar v }
  &
  \stackrel{\mathrm{Eq.\; \eqref{eq:vbar_def}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \sum_{i = 1}^n \rho_i \, \dot v_i
  %
  \stackrel{\mathrm{Eq.\; \eqref{eq:vt_diffeq_mbin}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  %
  \alpha(t)
  \sum_{j = 1}^n \frac{ h_j } { p_j }
                 \sum_{ i = 1 }^{n} \rho_i \, w_{ij}
  \\
  &
  \stackrel{\mathrm{Eq.\; \eqref{eq:w_balance}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t)
  \sum_{j = 1}^n \rho_j
    \, \frac{ h_j } { p_j }
  %
  \stackrel{\mathrm{Eq.\; \eqref{eq:w_sumj}}}
  {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t)
  \sum_{j = 1}^n w_{ij}
  \sum_{r = 1}^n \rho_r
    \, \frac{ h_r } { p_r }
  .
\end{aligned}
$$
}

Now by diagonalizing the matrix $\mathbf w$
using the eigenvectors defined in Eqs.
\eqref{eq:eig_w}-\eqref{eq:eig_orthonormal_rows},
we get a set of decoupled equations
%
\begin{equation}
  \dot y_k(t)
  =
  -\alpha(t) \, \lambda_k \, [y_k(t) - \theta_k(t)],
\label{eq:yt_diffeq}
\end{equation}
%
where
\begin{align}
  y_k(t)
  &=
  \sum_{j=1}^n
    \phi_{kj} \, x_j(t)
  ,
  \label{eq:y_def}
  \\
  \theta_k(t)
  &=
  \sum_{j=1}^n
    \phi_{kj} \, \xi_j(t)
  .
  \label{eq:eta_def}
\end{align}

@@

The error function defined in Eq. \eqref{eq:error_sum}
can be rewritten in terms of the eigenmodes
\begin{align}
  \Err
  \equiv
  \sum_{i = 1}^n \rho_i \,
    \left\langle x_i^2 \right\rangle
  =
  \sum_{i, j, k=1}^n \phi_{ki} \, \phi_{kj} \,
    \left\langle x_i \, x_j \right\rangle
  =
  \sum_{k = 1}^n
    \left\langle
      y_k^2
    \right\rangle
  ,
\notag
\end{align}
%
where we have used Eq. \eqref{eq:eig_orthonormal_cols}.
%
From Eq. \eqref{eq:eigenmode1},
the first eigenmode,
$y_1 = \sum_{j=1}^n \rho_j \, x_j$,
vanishes identically by Eq. \eqref{eq:rhox_sum}.
%
Thus, we can start the last sum %in Eq. \eqref{eq:y2_sum}
from $k = 2$,
\begin{equation}
  \Err
  =
  \sum_{k = 2}^n
    \left\langle
      y_k^2
    \right\rangle
  .
\label{eq:y2_sum}
\end{equation}

@@

    % xtics
    \draw[] (0.0, 0.1) -- (0.0, 0.0) node[below] {$0$};
    \draw[] (2.2, 0.1) -- (2.2, 0.0) node[below] {$q(T)$};
  \end{tikzpicture}
  %\makebox[\linewidth][c]{
  %  \includegraphics[angle=0, width=0.9\linewidth]{fig/massq.pdf}
  %}
  \caption{
    \label{fig:massq}
    Geometric characterization of the optimal schedule,
    $\alpha(t)$,
    via the normalized mass distribution, $m(Q)$.
    %
    \note{
      The figure was produced by \texttt{tikz}.
      % Gnuplot version by doc/fig/massq.gp
    }%
  }
\end{center}
\end{figure}

@@

Many commonly-used updating schemes
can be characterized by a rigid window
centered on the current bin.
%
We call these schemes homogeneous,
or translationally invariant.
%
An example is the Gaussian updating scheme
used in metadynamics.

By the translational invariance,
homogeneous updating schemes
satisfy detailed balance,
Eq. \eqref{eq:w_detailedbalance},
for the flat distribution, $\rho_i = 1/n$,
i.e., the updating matrix, $\mathbf w$,
is symmetric.
%
Besides, the updating matrices
share the same set of eigenvectors,
and thus are completely determined
by the eigenvalues.
%
Further, the multiplication of $\mathbf w$
can be reduced to a convolution with the window,
or the updating kernel.
%
Thus,
there is a one-to-one correspondence between
the eigenvalues of $\mathbf w$
and the updating kernel.

@@

\begin{table}[h]
  \caption{\label{tab:error_Gaussian}
    Error of the Gaussian updating scheme of $\sigma = 10$.
    $t_0$ is given by Eq. \eqref{eq:t0_sinc}.
    \note{
      The data were collected in
      \texttt{data/tab1} by running
      the program \texttt{invt}.
    }%
  }
  \setlength{\tabcolsep}{5pt}
  \begin{tabular} { l | p{2cm} p{2.5cm} p{1.8cm} }
    \hline
      Schedule
    &
      Inverse-time, Eq. \eqref{eq:alpha_invt1}
    &
      Modified inverse-time,
      Eq.~\eqref{eq:alpha_invtlambda}
      with optimized $\lambda$
    &
      Optimal,
      Eq. \eqref{eq:q_opt}
    \\
    \hline
    Global
    &
    $4.36(9) \times 10^{-6}$
    &
    $7.1(1) \times 10^{-7}$
    &
    $2.6(1) \times 10^{-7}$
    \\
    \hline
    Local
    &
    $3.83(7) \times 10^{-4}$
    &
    $1.72(4) \times 10^{-4}$
    &
    $9.1(2) \times 10^{-5}$
    \\
    \hline
  \end{tabular}
\end{table}

@@

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/xerr.pdf}
  }
  \caption{
    \label{fig:xerr}
    Error components for the Gaussian updating scheme
    with $\sigma = 10$
    from the optimal schedule
    and the inverse-time schedule, Eq. \eqref{eq:alpha_invt1}.
    %
    The latter schedule failed to damp out the errors
    from the long-wavelength (local) eigenmodes
    with $K \ge 5$.
    %
    For longer simulations,
    the error components from the optimal schedule
    would roughly maintain the same shape,
    but meet the initial error curve
    at a lower level (not shown).
    %
    \note{This figure was produced by \texttt{doc/fig/xerr.gp}.
      The data were saved under the directory \texttt{data/xerr}.
    }%
  }
\end{center}
\end{figure}

@@

Here, we evaluate the integrals of
the autocorrelation functions,
$\Gamma_k$'s,
in two special cases
in Secs. \ref{sec:Gamma_perfect}
and \ref{sec:Gamma_onestep}
in the asymptotic limit,
and describe a general method of measuring them
in Sec. \ref{sec:Gamma_measure}.

@@

Another special case is the
one-dimensional one-step process\cite{vankampen},
which models a local sampling process.
In each step, the system hops to
one of the two neighboring bins
with equal probability $1/2$.
%
The transition matrix is
%
\begin{equation}
  \arraycolsep=3.6pt\def\arraystretch{1.4}
  \mathbf A
  =
  \left(
    \begin{array}{cccccccc}
      \frac \sigma 2 & \frac 1 2 & 0 & \dots & \frac {1 - \sigma} 2 \\
      \frac 1 2 & 0         & \frac 1 2 & \dots & 0 \\
      \vdots & &  & & \vdots \\
      \frac {1 - \sigma} 2 & \dots &  & \frac 1 2 & \frac \sigma 2
    \end{array}
  \right),
\notag
%\label{eq:T_nn}
\end{equation}
%
where $\sigma = 0$ for a periodic variable, or $1$ otherwise.
%
For many updating schemes
(cf. Sec. \ref{sec:band-matrix}),
we can further assume that
the matrices $\mathbf A$ and $\mathbf w$
share the eigenvectors,
with $\rho_i = p_i = 1/n$.
%but not the eigenvalues.
%
The eigenvalues of $\mathbf A$ are
$$
  \Lambda_k
  =
  \cos \frac{ (k - 1) \, g \, \pi }
            {       2 \, n        }
  ,
$$
where
$g = 2$ for a periodic variable, or $1$ otherwise.
%Note that these $\Lambda_k$'s are
%not to be confused with the eigenvalues of $\mathbf w$,
%$\lambda_k$'s.
%
In this case,
the autocorrelation function $\kappa_k(t)$
decays as $\kappa_k(0) \, \Lambda_k^t$,
with $\kappa_k(0) = 1$ being the same as
the value in perfect sampling.
%
Thus,
%$\Gamma_k$ is roughly twice the autocorrelation time, and
from Eq. \eqref{eq:Gamma_sum}, we get
%
\begin{equation}
  \Gamma_k
  =
  1 + 2 \, \sum_{t = 1}^\infty \Lambda_k^t
  =
  \frac{ 1 + \Lambda_k }
       { 1 - \Lambda_k }
  =
  \cot^2 \left[
    \frac{ (k - 1) \, g \, \pi }
         { 2 \, n }
  \right]
  .
\label{eq:Gamma_onestep}
\end{equation}
